Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-10-25 manuscript/response_to_reviewers.txt

> In line 142 the method for the evaluation of the matrix effect is mentioned, but no 
> results are given.
The Boysen et al. (2018) paper details the method for identifying the magnitude of the 
matrix effect by using a full-strength and water-diluted pooled sample and we believe 
that the reader is best suited to the detailed analysis presented there than a
brief review of it in this manuscript. We have confirmed that references to
this method are cited immediately after its mention here.

> The experimental could be clearer in line 132 ff “The samples were then bead-beaten 
> three times, followed by triplicate extraction of the aqueous layer...” Was the 
> organic layer collected or was the aqueous layer collected (as suggested by the rest 
> of the sentence?). Was phase separation occurring spontaneously or was a 
> centrifugation required? Was separation done by pipetting?
We have added additional text to this line to clarify the requested points of confusion.

> Line 177: I would rather state “targeted analytical methods”. Metabolomics is by 
> definition rather untargeted.
While the exact definition of "targeted" vs "untargeted" metabolomics remains an 
active topic of debate, we believe that the methods used here are sufficiently
targeted (use of authentic standards and manual integration via Skyline/MSDIAL) to 
warrant the use of the term, especially in direct contrast to the automated/untargeted 
pipeline described immediately afterward and thus we have not made this change.

> Line 185: The data are discussed quantitatively but for some compounds no standard 
> was available. The claim that results were normalized “to the best match standard”. 
> How was this decided? Are the different response-factors in the MS considered?
As above, the Boysen et al. (2018) paper describes how internal standards are matched
to compounds for which no authentic standard was available based on the difference
in their response factor between a full-strength and half-strength pooled sample.
The only data presented quantitatively (i.e. as concentrations instead of relative
or absolute peak area) are those for which an authentic standard was available,
ensuring that each quantified compound's response factor was known. For untargeted
molecular features for which no authentic standard was available, data is only
discussed explicitly in peak area space or as a relative deviation between stations/
depths. We have also added links to the Github where the R code used to perform
these matching and normalization steps is available.

@@

> Chapter 4.1.1 It is strange to read an entire chapter on an outlier. It is really 
> one measurement that could be easily contaminated (e.g. one single zooplankton 
> organism on the filter would change everything) and I would not recommend to write 
> an entire chapter about the metabolome, risking to discuss artifacts and not data 
> important for the large picture. Is table 1 required.
We appreciate the reviewer's identification of this as an unhelpful distraction
from the main narrative and have removed chapter 4.1.1 entirely.

> Minor concerns:
> Could the CTD data be given in the supporting material, this would provide a view of 
> the eddies in a depth profiles.
The CTD data has been added to the Supplemental Tables in 3 forms - first, containing
a list of all the stations and their locations with the number of bottles fired. Second,
with all the associated metadata compiled from Barone 2022. Finally, the values used
in the manuscript's analysis after interpolation to the locations in space and time
from which metabolite samples were collected.

> Software: Give the version number of the software used (e.g. Proteowizard, msConvert, 
> XCMS...)
Done. Software versions for Proteowizard, Skyline, MSDIAL, R, XCMS, and vegan have been
added to the manuscript.

> Statistics: It would facilitate better interpretation if the statistical method used 
> for comparisons would be mentioned in the respective sections of results.
We have tried to add statistical methods used for comparisons in each section of the
results where applicable in this revision, largely annotating the places in which 
R-squared and p-values were obtained from PERMANOVAs. We have also added the sample
size to the Statistics section of the Results.

### 2024-10-25 manuscript/manuscript.Rmd
More stats details

Compounds for which we had an authentic standard (Supplemental Table 5) were manually integrated using Skyline (versions 4.1 and 21.2.0.470, @Adams2020) and MSDIAL (version 4.36, @Tsugawa2015) with the standard mixes used to ensure the correct peak was integrated by matching retention time. Compounds were removed from the analysis if they were visually assessed to have poor peak quality in the samples or peak areas in the samples similar to those in the methodological blanks. Skyline was used for both the eddy center samples and the HEE cruise data while MSDIAL was used for the eddy transect data due to the larger number of files in that dataset. Raw peak areas were then normalized to their best-matched internal standard (Supplemental Table 4) per @Boysen2018 except that all compounds were normalized to their best match no matter how minimal the improvement. Normalized peaks were then calibrated via a single-point standard curve as determined uniquely for each compound from the authentic standard mixes and used to convert normalized peak area into moles per µL injected and scaled to the amount of seawater filtered to provide a final estimate of environmental concentration for each compound in moles per liter of seawater filtered.

@@

Samples were also processed via the untargeted workflow detailed in @Kumler2023. In short, XCMS (version 3.10.0, @R-xcms) was used for feature detection with CentWave peakpicking, Obiwarp alignment, and peak density correspondence/grouping. This was followed by manual inspection of the extracted ion chromatogram for each mass feature by an expert for quality with only those peaks classified as "Good" used in the analysis here. Features were then matched with and normalized to an internal standard (Supplemental Table 1) as described above.

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths ($\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = 0.813, p-value < 0.001, Supplemental Figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental Figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= 0.593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= 0.503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= 0.049$, p-value = 0.792). This meant that metabolites contributed a relatively fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 3). Particulate nitrogen was tightly correlated with particulate carbon ($r=0.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.

>>>>

Compounds for which we had an authentic standard (Supplemental Table 5) were manually integrated using Skyline (versions 4.1 and 21.2.0.470, @Adams2020) and MSDIAL (version 4.36, @Tsugawa2015) with the standard mixes used to ensure the correct peak was integrated by matching retention time. Compounds were removed from the analysis if they were visually assessed to have poor peak quality in the samples or peak areas in the methodological blanks similar to those in the samples. Skyline was used for both the eddy center samples and the HEE cruise data while MSDIAL was used for the eddy transect data due to the larger number of files in that dataset. Raw peak areas were then normalized to their best-matched internal standard (Supplemental Table 4) per @Boysen2018 except that all compounds were normalized to their best match no matter how minimal the improvement. While significant analytical drift was detected during the longer eddy transect sample run, the use of internal standards was able to negate the instrument's general increase in sensitivity. Normalized peaks were then calibrated via a single-point standard curve as determined uniquely for each compound from the authentic standard mixes and used to convert normalized peak area into moles per µL injected and scaled to the amount of seawater filtered to provide a final estimate of environmental concentration for each compound in moles per liter of seawater filtered. Scripts used for this analysis are available at https://github.com/wkumler/MesoscopeMetabolomicsManuscript/tree/master/untargeted.

@@

Samples were also processed via the untargeted workflow detailed in @Kumler2023. In short, XCMS (version 3.10.0, @R-xcms) was used for feature detection with CentWave peakpicking, Obiwarp alignment, and peak density correspondence/grouping. This was followed by manual inspection of the extracted ion chromatogram for each mass feature by an expert for quality with only those peaks classified as "Good" used in the analysis here. Features were then matched with and normalized to an internal standard (Supplemental Table 1) as described above. Scripts used for this analysis are available at https://github.com/wkumler/MesoscopeMetabolomicsManuscript/tree/master/untargeted.

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths (Type I linear regression $\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = 0.813, p-value < 0.001, Supplemental Figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental Figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= 0.593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= 0.503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= 0.049$, p-value = 0.792). This meant that metabolites contributed a relatively fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 3). Particulate nitrogen was tightly correlated with particulate carbon ($r=0.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.
